---
title: "5. Hands-on: missing outcomes"
date-modified: last-modified
execute:
  warning: false
  message: false
---

```{r setup, include = FALSE}
# Render-time guard: if TwoStageTMLE isn't installed, fall back to
# eval: false so the rest of the site still builds for reviewers.
have_pkg <- requireNamespace("TwoStageTMLE", quietly = TRUE)
knitr::opts_chunk$set(eval = have_pkg, collapse = TRUE, comment = "#>")
```

```{r install-msg, eval = !have_pkg, echo = FALSE, results = "asis"}
cat("::: {.callout-warning}\n",
    "The `TwoStageTMLE` package is not installed; code below is shown but not run. ",
    "See the [README](README.md) for installation steps.\n",
    ":::\n", sep = "")
```

::: {.callout-note .draft}
**DISCLAIMER: This page is under construction**  
:::

::: {.callout-tip}
**Live R in the browser.** This site uses [webR](https://docs.r-wasm.org/webr/).
Chunks marked `{webr-r}` are editable — change the code, click *Run*, see the
new output. No R install needed. (The cells that call `TwoStageTMLE` directly
still render server-side until the package ships a WebAssembly binary; see
the README's roadmap.)
:::

This chapter walks through the **foundational** Two-Stage TMLE pipeline
[@balzer2021twostage] on a simulated CRT with **missing outcomes**. We:

1. Generate one trial — 30 clusters, 100–200 participants each, pair-matched.
2. Run **Stage 1** within each cluster (TMLE for the cluster-specific mean
   outcome, adjusting for differential measurement).
3. Run **Stage 2** at the cluster level with Adaptive Pre-specification.
4. Report the point estimate, 95% CI, and inspect the chosen adjustment
   variable.

## Generate one CRT

```{r generate}
library(TwoStageTMLE)
library(data.table)

set.seed(1)
J  <- 30
n  <- c(100, 150, 200)
X  <- TwoStageTMLE:::gen_data_complex(effect = TRUE, n = n, J = J)
O  <- TwoStageTMLE:::get_obs(X, exclude = FALSE)

head(O[, c("id", "pair", "A", "X1", "X2", "M", "Delta", "Y")])
```

`Delta` indicates outcome measurement (0 if missed); `Y` is `NA` where
`Delta = 0`. `X1`, `X2`, `M` are individual covariates; `X1.c`, `X2.c`
are the cluster-level means.

::: {.callout-note}
**Try it.** This cell runs **in your browser**. Tweak the parameters and
see how the cluster-level summaries shift before any TMLE has been
applied — useful intuition for why Stage 1 matters.
:::

```{webr-r}
#| label: explore-missingness-in-browser
#| autorun: false
library(data.table)
set.seed(1)

# Mini CRT: J clusters, n_per individuals per cluster.
J <- 10; n_per <- 80

dat <- rbindlist(lapply(seq_len(J), function(j) {
  W      <- rnorm(n_per)
  A      <- rbinom(1, 1, 0.5)                            # cluster trt
  pDelta <- plogis(-0.3 + 0.7*A + 0.6*W)                 # measurement depends on W and A!
  Delta  <- rbinom(n_per, 1, pDelta)
  Y      <- rbinom(n_per, 1, plogis(-0.5 + 0.5*A + 0.8*W))
  Y[Delta == 0] <- NA_integer_
  data.table(id = j, A = A, W = W, Delta = Delta, Y = Y)
}))

# Naive complete-case mean per cluster (biased) vs. an honest fraction-measured.
dat[, .(meanY_naive = mean(Y, na.rm = TRUE),
        pct_measured = mean(Delta)),
    by = .(id, A)][order(A, id)]
```


## Stage 1 — within-cluster missing-data adjustment

```{r stage1}
stage1 <- stage1_tmle(
  data       = O,
  ind_cov    = c("X1", "X2", "M"),
  clust_cov  = c("X1.c", "X2.c"),
  sl_library = c("SL.mean", "SL.glm", "SL.gam")
)
stage1[, .(id, pair, A, X1.c, X2.c, unadj, tmle)]
```

`unadj` is the simple mean among the measured; `tmle` is the cluster
mean adjusted for differential measurement.

## Stage 2 — cluster-level TMLE + Adaptive Pre-specification

```{r stage2}
s2 <- copy(stage1)
s2[, Y := tmle]
s2 <- add_weights(s2, target = "clust")

cand <- cand_adj(c("X1.c", "X2.c"))

est <- stage2_tmle(
  data          = s2,
  goal          = "RD",
  target        = "clust",
  break_match   = FALSE,           # preserve the matched pairs
  do_data_adapt = TRUE,
  cand_q_adj    = cand$cand_q_adj, cand_q_form = cand$cand_q_form,
  cand_g_adj    = cand$cand_g_adj, cand_g_form = cand$cand_g_form,
  remove_pscore = TRUE
)
est[, .(est, ci_lo, ci_hi, se, pval, q_adj, g_adj)]
```

## What APS chose

`q_adj` and `g_adj` index the candidate adjustment vector — `1 = "U"`
(the unadjusted estimator), `2 = "X1.c"`, `3 = "X2.c"`. With small
clusters and well-balanced randomization, APS often picks the unadjusted
estimator for the propensity score (because the trial-arm assignment is
known to be 0.5) and the most-predictive covariate for the outcome.

## End-to-end driver

```{r e2e, eval = FALSE}
out <- two_stage_tmle(
  data       = O,
  ind_cov    = c("X1", "X2", "M"),
  clust_cov  = c("X1.c", "X2.c"),
  goal       = "RD", target = "clust", break_match = FALSE,
  do_data_adapt = TRUE,
  cand_q_adj = cand$cand_q_adj, cand_q_form = cand$cand_q_form,
  cand_g_adj = cand$cand_g_adj, cand_g_form = cand$cand_g_form,
  remove_pscore = TRUE
)
out$est        # the Stage 2 estimate
out$stage1     # the cluster-level table from Stage 1
```

## Where to go next

- [Chapter 6](06_R_mediated.qmd) — when the missing variable is
  *eligibility* (mediated by screening), not the outcome.
- The package vignette `vignette("reproduce-paper")` walks through the
  full simulation grid of @balzer2021twostage.

## Resources

::: {.callout-tip}
**Reference cards**

- `stage1_tmle()` documentation: `?TwoStageTMLE::stage1_tmle`
- `stage2_tmle()` documentation: `?TwoStageTMLE::stage2_tmle`
- `cand_adj()` documentation: `?TwoStageTMLE::cand_adj`
:::
