---
title: "6. Hands-on: the multilevel-mediation-missing problem"
date-modified: last-modified
execute:
  warning: false
  message: false
---

```{r setup, include = FALSE}
have_pkg <- requireNamespace("TwoStageTMLE", quietly = TRUE)
knitr::opts_chunk$set(eval = have_pkg, collapse = TRUE, comment = "#>")
```

```{r install-msg, eval = !have_pkg, echo = FALSE, results = "asis"}
cat("::: {.callout-warning}\n",
    "The `TwoStageTMLE` package is not installed; code below is shown but not run. ",
    "See the [README](README.md) for installation steps.\n",
    ":::\n", sep = "")
```

This chapter walks through the **mediated** Stage 1 estimator
[@nakato2026mediated] on a simulated CRT where:

* The intervention $A$ aims to expand both **risk screening** and **uptake
  of biomedical prevention**.
* The eligibility variable $Y_1^*$ (e.g., true HIV risk status) is
  observed *only when screened* (when $M = 1$).
* The outcome $Y_2$ (e.g., PrEP uptake) is fully observed (e.g., clinical
  records).

The cluster-level endpoint is the **ratio**:

$$
Y^c \;=\; \frac{\mathbb{P}(Y_2 = 1)}{\mathbb{E}\!\bigl[\mathbb{P}(Y_1 = 1 \mid M = 1, W)\bigr]}.
$$

Numerator: empirical proportion uptaking. Denominator: TMLE for the
prevalence of risk, fit within each cluster.

## Generate one CRT

We use a condensed version of the Nakato (2026) DGP. The intervention
boosts screening; outcomes only happen among the screened-and-eligible.

```{r dgp}
library(TwoStageTMLE)
library(data.table)

set.seed(2026)

gen_cluster <- function(N, j) {
  UE1 <- rep(runif(1, -1, 1), N); UE2 <- rep(runif(1, -1, 1), N)
  E1  <- rep(rnorm(1, UE1[1], 0.5), N); E2 <- rep(rnorm(1, UE2[1], 0.5), N)
  W1  <- rnorm(N, 0, 0.5); W2  <- rnorm(N, 0, 0.5); W3 <- rbinom(N, 1, 0.5)
  pY1 <- plogis(0.5 + W1 + W2 - W3 + 0.25*E1 + 0.25*E2)
  data.table(
    id = j, E1 = E1, E2 = E2, W1 = W1, W2 = W2, W3 = W3,
    UE1 = UE1, UE2 = UE2, Y1_star = as.integer(runif(N) < pY1)
  )
}

J <- 20
dat <- rbindlist(lapply(seq_len(J), function(j) gen_cluster(120, j)))
A_per <- sample(c(rep(1L, J/2), rep(0L, J/2)))
dat[, A := A_per[id]]

pM <- plogis(-0.5 + 0.6*dat$A + 0.5*dat$W1 + 0.4*dat$W2 -
              0.4*dat$A*dat$W3 + 0.4*(1-dat$A)*dat$W3 +
              0.1*dat$UE1 + 0.1*dat$UE2)
dat[, M  := as.integer(runif(.N) < pM)]
dat[, Y1 := Y1_star * M]

pY2 <- plogis(0.2 + 0.1*dat$A + dat$W1 + 0.5*dat$W2 +
              2*dat$W3 + 0.2*dat$E1 + 0.2*dat$E2)
dat[, Y2 := ifelse(Y1_star == 1 & M == 1,
                   as.integer(runif(.N) < pY2),
                   0L)]

dat[, c("UE1", "UE2", "Y1_star") := NULL]
head(dat[, .(id, A, W1, W2, W3, E1, E2, M, Y1, Y2)])
```

## Stage 1 — ratio endpoint per cluster

```{r stage1}
s1 <- stage1_mediated(
  data       = dat,
  ind_cov    = c("W1", "W2", "W3"),
  clust_cov  = c("E1", "E2"),
  sl_library = c("SL.mean", "SL.glm")
)
s1[, .(id, A, E1, E2, screened, eligible, unadj, tmle)]
```

Each row is one cluster. The four candidate endpoints:

| Column     | Estimand                              | Bias               |
|------------|---------------------------------------|--------------------|
| `screened` | $\mathbb{P}(Y_2 \mid M = 1)$                  | Biased (conditions on a post-treatment mediator) |
| `eligible` | $\mathbb{P}(Y_2 \mid Y_1 = 1)$                | Biased (conditions on observed-eligible) |
| `unadj`    | $\mathbb{P}(Y_2) / \mathbb{P}(Y_1 \mid M = 1)$        | Unadjusted denominator |
| **`tmle`** | $\mathbb{P}(Y_2) / \mathrm{TMLE}[\mathbb{P}(Y_1 \mid M = 1, W)]$ | **Recommended endpoint** |

The first two exist so you can see how conditioning the wrong way shifts
the estimate; `tmle` is what gets fed to Stage 2.

## Stage 2 — cluster-level TMLE on the ratio endpoint

Stage 2 is *unchanged* from [Chapter 5](05_R_foundational.qmd) — the
ratio endpoint slots into the same machinery:

```{r stage2}
s2 <- copy(s1)
s2[, Y := tmle]
s2 <- add_weights(s2, target = "clust")

cand <- cand_adj(c("E1", "E2"))

est <- stage2_tmle(
  data          = s2,
  goal          = "RD",
  target        = "clust",
  break_match   = TRUE,
  do_data_adapt = TRUE,
  cand_q_adj    = cand$cand_q_adj, cand_q_form = cand$cand_q_form,
  cand_g_adj    = cand$cand_g_adj, cand_g_form = cand$cand_g_form,
  remove_pscore = TRUE
)
est[, .(est, ci_lo, ci_hi, se, pval, q_adj, g_adj)]
```

## End-to-end driver

```{r e2e, eval = FALSE}
out <- two_stage_tmle_mediated(
  data          = dat,
  ind_cov       = c("W1", "W2", "W3"),
  clust_cov     = c("E1", "E2"),
  sl_library    = c("SL.mean", "SL.glm"),
  goal          = "RD", target = "clust", break_match = TRUE,
  do_data_adapt = TRUE,
  cand_q_adj    = cand$cand_q_adj, cand_q_form = cand$cand_q_form,
  cand_g_adj    = cand$cand_g_adj, cand_g_form = cand$cand_g_form,
  remove_pscore = TRUE
)
out$est        # Stage 2 estimate
out$stage1     # cluster table with all four candidate endpoints
```

## A note on variance

Stage 2 computes its variance from the cluster-level influence curve and
treats the Stage 1 ratio as a *fixed observation at the cluster level*.
The Stage-1 contribution to variance is asymptotically negligible when
within-cluster sample sizes are large. Nakato (2026) verifies acceptable
coverage at $N_c \approx 200$ per cluster, $J \in \{20, 30, 50, 70\}$.

For inference also robust to the Adaptive Pre-specification selection,
pass `do_cv_variance = TRUE` to `stage2_tmle()` — the returned data.table
gets additional `cv_*` columns with the CV CI and p-value.

## Where to go next

- The package vignette `vignette("mediated-missing")` covers the same
  flow with additional commentary.
- [Chapter 7](11_info_conclusion.qmd) closes with the next steps — Aim
  1B (effect heterogeneity), Aim 2 (SAPPHIRE & Peer-led DCP secondary
  analyses), and the Roadmap Shiny app.

## Resources

::: {.callout-tip}
**Reference cards**

- `stage1_mediated()` documentation: `?TwoStageTMLE::stage1_mediated`
- `two_stage_tmle_mediated()` documentation: `?TwoStageTMLE::two_stage_tmle_mediated`
- Nakato & Balzer (2026) — manuscript [@nakato2026mediated].
:::
